---
title: "Computational Statistics Lab 1"
author: "Damian Ke & Kyriakos Papadopoulos"
date: "2022-14-11"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Libraries

```{r, results='hide', warning=FALSE}
library(ggplot2)
```

## Question 1: Be careful when comparing

**Q.1**: Check the results of the snippets. Comment what is going on.

```{r}
x1 <-  1/3
x2 <-  1/4
if(x1 - x2==1/12){
  print("Subtraction is correct")
  }else{
  print("Subtraction is wrong")
}

x1 <- 1 
x2 <- 1/2
if(x1 - x2==1/2){
  print("Subtraction is correct")
  }else{
  print("Subtraction is wrong")
}

```

**A**: The first code snippet is mathematically correct, but due to
underflow the program is incorrect. It is because $\frac{1}{3}$. has
infinitive values after the decimal point and computers have limited
storage value (regarding to type of variable, which in our case is a
double and in R it is represented in 53 bits) and cannot represent all
digits of $\frac{1}{3}$ .

The second code snippet is correct, because $\frac{1}{2}$ is exactly 0.5
which does not have infinitive values after decimal point and can be
represented exactly in R.

**Q.2**: If there are any problems, suggest improvements.

**A**: Is a utility to compare R objects x and y testing 'near
equality'. If they are different, comparison is still made to some
extent, and a report of the differences is returned

The all.equal() function compares two objects at a "near equality",
which means that if difference between the values is very small then it
considers that the values are equal and returns true. \$a_1 - e \< a_2
\< a_1 + e \$ Where e is a very small number.

```{r}
x1 <-  1/3
x2 <-  1/4
#1.2
if(isTRUE(all.equal(x1-x2,1/12))){
  print("Subtraction is correct")
}else{
  print("Subtraction is wrong")
}
```

## Question 2: Derivative

**Q.1**: Write your own R function to calculate the derivative of
$f(x) = x$ in this way with $\epsilon=10^{-15}$.

**A**:

```{r}
derivative <- function(x){
  out =(x + 10^-15 - x)/10^-15
  return(out)
}
```

**Q.2**: Evaluate your derivative function at $x=1$ and $x = 100000$.

**A**:

```{r}
print(derivative(1))
print(derivative(100000))
```

**Q.3**: What values did you obtain? What are the true values? Explain
the reasons behind the discovered differences.

**A**: We got $1.110223$ and $0$, the real values should be $1$ as the x
cancel each other and the $\frac{\epsilon}{\epsilon} = 1$. The issue is because 
of underflow in both cases. In the first case, we have rounding error because R can not represent x + 10^-15 as it returns 1.000000000000001110223.
In the second case, in the nominator $100000 + \epsilon$ gives $100000$ as $\epsilon$ 
is very small and can't be added correctly to 100000. In general in both cases we have precision problems as we loose significant digits



## Question 3: Variance

**Q.1**: Write your own R function, myvar, to estimate the variance in
this way

**A**:

```{r}
myvar <- function(x){
  n = length(x)
  result = ((1/(n-1))*(sum(x^2)- (1/n)*(sum(x)^2)))
  return (result)
}
```

**Q.2**: Generate a vector $x= (x_1,...,x_{10000})$ with 10000 random
numbers with mean $10^8$ and variance 1.

**A**:

```{r}
t = rnorm(10000, 10^8, 1)
```

**Q.3**: For each subset \$ X_i = {x_1,...,x_i}, i = 1,...,10000\$
compute the difference $Y_i = myvar(X_i)-var(X_i)$, where $var(X_i)$ is
the standard variance estimation function in R. Plot the dependence
$Y_i$ on $i$. Draw conclusions from this plot. How well does your
function work? Can you explain the behaviour?

**A**

```{r}
test = c()
for(i in 1:length(t)){
  a = myvar(t[1:i]) - var(t[1:i])
  test = append(test, a)
}

data = data.frame("Values" = test)
data["i"] = 1:length(t)
data[1,1] <- 0

p <- ggplot(data, aes(x=i, y=as.numeric(Values))) +
  geom_point() +
  ylab("Error")+
  ggtitle("Scatterplot, i over Error")
p
```


The variance in t is equal to 1 so most of the values in the plot are -1 which means that we get as a result 0 instead of 1. This is can easily explained as we have to add  values of similar magnitude but of opposite sign so it leads to catastrophic cancellation and it gives zero.

**Q4**: How can you better implement a variance estimator? Find and
implement a formula that will give the same results as var()

```{r}

better_myvar <- function(x){
  sum((x- mean(x))^2) / (length(x) - 1)
}

test2 = c()
for(i in 1:length(t)){
  a = better_myvar(t[1:i]) - var(t[1:i])
  test2 = append(test2, a)
}

data = data.frame("Values" = test2)
data["i"] = 1:length(t)
data[1,1] <- 0

p2 <- ggplot(data, aes(x=i, y=as.numeric(Values))) +
  geom_point()+
  ylab("Error")+
  ggtitle("Scatterplot, i over Error")
p2
```

**A**: The variance method can be improved by using normal formula for
varience. $\frac{(\sum_{i = 1}^{n} X_i-\overline{X})^2}{n-1}$. As it can
be seen in the plot, the error is very small around 0 and it can be
considered as sucessful method.

## Question 4: Binomial coeffcient

**Q.1**: Even if overflow and underfl ow would not occur these
expressions will not work correctly for all values of n and k. Explain
what is the problem in A, B and C respectively.

```{r}
bf1 <- function(n, k){
  out <- prod ( 1 : n ) / (prod ( 1 : k ) * prod ( 1 : ( n - k ) ) )
  return(out)
}

bf2 <- function(n, k){
  out <- prod( ( k+1) : n ) / prod ( 1 : ( n - k ) )
  return(out)
}

bf3 <- function(n, k){
  out <- prod ( ( ( k+1) : n ) / ( 1 : ( n - k ) ) )
  return(out)
}

```

**A**: If n = k then the demoninator will be 0 so the value that it will
return is infinitive in all of 3 cases.

**Q.2**: In mathematical formula one should suspect overflow to occur
when parameters, here n and k, are large. Experiment numerically with
the code of A, B and C, for different values of n and k to see whether
overflow occurs. Graphically present the results of your experiments.

```{r}
n_values = c(0:1000)
k_values = c(0:1000)
#results_list = dataframe(c=("N", ))
results1 = c()
results2 = c()
results3 = c()
n_values_list = c()
k_values_list = c()
x = 1
for(i in k_values){
  for(j in n_values){
  if (i <= j){  
    results1[x] <- bf1(j,i)
    results2[x] <- bf2(j,i)
    results3[x] <- bf3(j,i)
    k_values_list[x] <- i
    n_values_list[x] <- j
    x=x+1
  }
  }
}


frame = data.frame(results1, results2, results3, n_values_list, k_values_list)
colnames(frame) = c("df1", "df2", "df3", "N","K")

frame["results1"] <- ifelse((is.na(frame["df1"])),"Not a Value","Value")

p_4 <-  ggplot(frame, aes(x = N, y = K, fill = as.factor(results1))) +
  geom_tile()+
  labs(fill="Definition of observation")+
  ggtitle("Heatmap of Method(A)")
p_4

frame["results2"] <- ifelse((is.na(frame["df2"])),"Not a Value","Value")

p_5 <-  ggplot(frame, aes(x = N, y = K, fill = as.factor(results2))) +
  geom_tile()+
  labs(fill="Definition of observation")+
  ggtitle("Heatmap of Method(B)")
p_5

frame["results3"] <- ifelse((is.na(frame["df3"])),"Not a Value","Value")

p_6 <-  ggplot(frame, aes(x = N, y = K, fill = as.factor(results3))) +
  geom_tile()+
  labs(fill="Definition of observation")+
  ggtitle("Heatmap of Method(C)")
p_6

```

**Q3**: Which of the three expressions have the overflow problem?
Explain why

**A**: The first 2 have the overflow problem because they consist of 2
prod one in the numerator and one in the denominator.In the first case when k is 
too big n-k as well, the output will be $\frac{Inf}{Inf}$ as the product of a very 
big sequence is *Inf* in R. When it happens that in R it returns *NaN* value.

In the second case we have *NaN* when only n-k is a very big number because The 
output will be $\frac{Inf}{Inf}$. As mentioned above R in that case returns *NaN*


So in the 3rd case it will output *Inf* as both numerator and denominator are 
inside the same product and it will return Inf for very big numbers and never *NaN*

