---
title: "Assignment2"
author: "Damian Ke & Kyriakos Papadopoulos"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Gibbs sampling

## Question 1
What kind of model is reasonable to use here?
```{r, warning=FALSE}
load("chemical.RData")

library(ggplot2)

data= data.frame(X,Y)
ggplot(data, aes(x=X,y=Y))+
  geom_point()
```

**Answer**
As there seem to be a correlation between X and Y. A linear or polynomial regression can be a reasonable
model to use here.

## Question 2
Present the formulae showing the likelihood $p(\overrightarrow{Y}\mid\overrightarrow{\mu})$ and the prior $p(\overrightarrow{\mu})$

**Answer**

$$p(\overrightarrow{\mu}) = 1* N(\mu_1,0.2)*N(\mu_2,0.2)*N(\mu_3,0.2)*...*N(\mu_{n-1},0.2)$$
Which is equal to $p(\overrightarrow{\mu}) = (\frac{1}{0.2*\sqrt{2*\pi}})^{n-1}*e^{(-\frac{\sum_{i=2}^{n}(\mu_{i}-\mu_{i-1})^2}{0.08})}$

$$p(\overrightarrow{Y}\mid\overrightarrow{\mu})= (\frac{1}{0.2*\sqrt{2*\pi}})^n*e^{(-\frac{\sum_{i=1}^{n}({y_i}-\mu_{i})^2}{0.08})} $$ 






# Question 3
Use Bayes' Theorem to get the posterior up to a constant proportionality, and then find
out the distributions of $p({\mu_i}\mid\overrightarrow\mu_{-i},\overrightarrow{Y})$, where $\mu_{-i}$ is a vector containing all $\mu$ values except of $\mu_i$

**Answer**
To get posterior 
$$p({\overrightarrow\mu}\mid\overrightarrow{Y}) \propto p(\overrightarrow{Y}\mid\overrightarrow{\mu}) * p(\overrightarrow{\mu})$$

It is also important to mention that
$p({\overrightarrow\mu},\overrightarrow{Y}) = p(\overrightarrow{Y}\mid\overrightarrow{\mu}) * p(\overrightarrow{\mu})$ as they are dependent on each other.

Therefore, $p({\mu_n}\mid\overrightarrow{\mu}_{-n},\overrightarrow{Y}) = \frac{p(\overrightarrow{\mu}_{-n},\overrightarrow{Y},\mu_n)}{p(\overrightarrow{\mu}_{-n},\overrightarrow{Y})}$
and $p(\overrightarrow{\mu}_{-n},\overrightarrow{Y},\mu_n) = p({\overrightarrow\mu},\overrightarrow{Y})$.

It can therefore be concluded that $(\frac{1}{0.2*\sqrt{2*\pi}})$ will be same for all the cases.
Therefore, the focus is on the exponents.????????

$$p(\overrightarrow{\mu}\mid\overrightarrow{Y}) \propto e^{(-\frac{\sum_{i=1}^{n}({y_i}-\mu_{i})^2}{0.08})}* e^{(-\frac{\sum_{i=1}^{n-1}({\mu_{i+1}}-\mu_{i})^2}{0.08})}$$
Multiplication of these two negative exponents moves places of (a-x) to (x-a).

Due to $\frac{p(\overrightarrow{\mu}_{-n},\overrightarrow{Y},\mu_n)}{p(\overrightarrow{\mu}_{-n},\overrightarrow{Y})}$ The $\Sigma$ will be canceled out as the only part being left is $\mu_n$.

For case when 1:
$p(\mu_1\mid\overrightarrow{\mu_{-1}},\overrightarrow{Y})\propto e^{(-\frac{((\mu_1-y_1)^2+(\mu_1-\mu_2)^2)}{0.08})}$
and through Hint B
$$e^{(-\frac{({\mu_1}-(y_1+\mu_{2})/2)^2+}{0.08/2})} \sim N(\frac{y_i+\mu_2}{2},0.04)$$


For case when n:
$p(\mu_n\mid\overrightarrow\mu_{-n},\overrightarrow{Y}) \propto e^{(-\frac{(\mu_{n}-{y_n})^2+(\mu_n-\mu_{n-1})^2}{0.08})}$ 
and through Hint B
$$e^{(-\frac{({\mu_n}-(y_n+\mu_{n-1})/2)^2+}{0.08/2})} \sim N(\frac{y_n+\mu_{n-1}}{2},0.04)$$


For cases when i:
The exponent from the prior has additional case for i as there is a previous and next step.
$p({\mu_i}\mid\overrightarrow\mu_{-i},\overrightarrow{Y}) \propto e^{(-\frac{({y_i}-\mu_{i})^2+(\mu_i-\mu_{i-1})^2+(\mu_{i}-\mu_{i+1})^2}{0.08})}$ 
and through Hint C
$$e^{(-\frac{({\mu_i}-(y_i+\mu_{i-1}+\mu_{i+1})/3)^2+}{0.08/3})} \sim N(\frac{y_i+\mu_{i-1}+\mu_{i+1}}{2},0.08/3)$$

## Question 4
Plot the expected value
of $\overrightarrow\mu$ versus X and Y versus X in the same graph. Does it seem that you have managed to
remove the noise? Does it seem that the expected value of $\overrightarrow\mu$ can catch the true underlying
dependence between Y and X?

```{r, warning=FALSE}

Gibbs<-function(nstep,vmean,mVar){
  n<-length(vmean)
  mX<-matrix(0,nrow=nstep,ncol=n)
  #Starting point
  mX[1,] = rep(0,n)
  
  #Loop to get 1000 values
  for (x in 2:nstep){
    #Case i = 1
    mX[x,1] = rnorm(1,mean=(vmean[1]+mX[x-1,2])/2,sd=sqrt((2*mVar**2)/2))
    #For case i
    for (i in 2:(n-1)){
      mX[x,i] = rnorm(1,mean=(vmean[i]+mX[x,i-1]+mX[x-1,i+1])/3,sd=sqrt((2*mVar**2)/3))
    }
    #Case i = n
    mX[x,n] = rnorm(1,mean=(vmean[n]+mX[x,n-1])/2,sd=sqrt((2*mVar**2)/2))
  }
  mX
}

set.seed(12345)

results<-Gibbs(1000,Y,0.2)

library(ggplot2)
mean_results = colMeans(results)
data["mean_results"] = mean_results

p = ggplot()+
  geom_line(aes(x=X,y=mean_results), color="red")+
  geom_line(aes(x=X,y=Y),color="blue")+
  ylab("Y")
p
```

There seem to be no noise as the line corresponding for $\overrightarrow\mu$ is smooth. 
It seems that the expected value of $\overrightarrow\mu$ catches the true underlying dependence between Y and X as both of the lines are having almost the same values..


## Question 5
Comment on the burn-in period and convergence.
```{r}
results_df = data.frame(1:1000,results[,50])
colnames(results_df) = c("ID", "Results")
p2 = ggplot(results_df,aes(x=ID,y=Results))+
  geom_line()+
  xlab("nstep")
p2


```

**Answer**
For the beginning of the trace plot, there exist a small burn-in period.
Thereafter, the plot converges.
