---
title: "Assignment 2"
author: "Damian Ke and Kyriakos Papadopoulos"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 2: Bootstrap, jackknife and confidence intervals

# Question 1
Plot the histogram of Price. Does it remind any conventional distribution? Compute the
mean price.
```{r, echo=FALSE}
library(ggplot2)
df = read.csv2("prices1.csv")
p = ggplot(df, aes(x=Price)) +
  geom_histogram(binwidth=150)
p
mean_price = mean(df$Price)
```

The mean price is equal to `r mean_price`

**Answer**
The distribution looks like gamma distribution.

# Question 2
Estimate the distribution of the mean price of the house using bootstrap. Determine the
bootstrap bias–correction and the variance of the mean price. Compute a 95% confidence
interval for the mean price using bootstrap percentile, bootstrap BCa, and first–order
normal approximation.

```{r, echo=FALSE, warning=FALSE}
library(boot)
set.seed(12345)
stat1 = function(data,vn){
  mean(data[vn,]$Price)
}
res=boot(df,stat1,R=1000)
B=1000
t_1 = 2*res$t0-mean(res$t)
var_t = (1/(B-1))*(sum((res$t-mean(res$t))^2))
```
**Clarification** Used formulas are accordingly to the lecture slides.

The mean price from bootstrap is equal to `r res$t0`

95% Confidence interval with bootstrap percentile is equal to: (`r boot.ci(res)$percent[4:5]`)

95% Confidence interval with bootstrap BCa is equal to: (`r boot.ci(res)$bca[4:5]`)

95% Confidence interval with bootstrap  first–order
normal approximation is equal to: `r boot.ci(res)$normal[1,2:3]`

Bootstrap bias–correction is equal to: `r t_1`

Bootstrap variance is equal to: `r var_t`

# Question 3
Estimate the variance of the mean price using the jackknife and compare it with the
bootstrap estimate

```{r, echo=FALSE}
n = length(df$Price)
t_j = rep(0,n)
for (j in 1:n){
  t_j[j] = n*mean(df$Price)-(n-1)*mean(df$Price[-j])
}
j_t = mean(t_j)
var_t2 = (1/(n*(n-1)))*sum((t_j-j_t)**2)
```

**Clarification**: n is set equal to the length of the data sample
as according to the course literature when k = 1. 
According to the lecture slides B=n. By testing both options, there is a small difference
between the values.

Jackknife variance is equal to: `r var_t2`
Bootstrap variance is equal to: `r var_t`

**Answer**
The variance of the mean of jackknife is larger than the bootstrap variance of the mean.
The jackknife leaves one of the observations and then calculates the variance
of the new subset for n amount of iterations. Which compared to bootstrap variance of the mean
will overestimate the variance.


## Question 4
Compare the confidence intervals obtained with respect to their length and the location of
the estimated mean in these intervals.
```{r, echo=FALSE, warning=FALSE}
interval = boot.ci(res)
interval_percentile = interval$percent[1,4:5]
interval_bca = interval$bca[1,4:5]
interval_normal = interval$normal[1,2:3]

interval_percentile[3] = interval_percentile[2]-interval_percentile[1]
interval_percentile[4] = (interval_percentile[2]-interval_percentile[1])/2
interval_bca[3] = interval_bca[2]-interval_bca[1]
interval_bca[4] = (interval_bca[2]-interval_bca[1])/2
interval_normal[3] = interval_normal[2]-interval_normal[1]
interval_normal[4] = (interval_normal[2]-interval_normal[1])/2

df_results = t(data.frame(interval_bca,interval_normal,interval_percentile))
colnames(df_results) = c("Lower CI", "Upper CI", "Length","Mean")
rownames(df_results) = c("BCA", "Normal","Percentile")

knitr::kable(df_results)
```

**Answer**
As it can be seen in the table, BCa has the largest length difference.
Thereafter it is Percentile and lastly Normal(first–order normal approximation). 
The mean for corresponding intervals are in order of BCA as largest then Percentile and 
lastly Normal (first–order normal approximation)




