---
title: "Assignment2"
author: "Damian Ke and Kyriakos Papadopoulos"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Question 2: EM algorithm

## Question 2.1
Make a time series plot describing dependence of Z and Y versus X. Does it seem that two
processes are related to each other? What can you say about the variation of the response
values with respect to X?

```{r,echo=FALSE}
library(ggplot2)
df= read.csv("physical1.csv")

#Question 1
ggplot(data=df,aes(x=X))+
  geom_line(aes(y=Z, color="Z"))+
  geom_line(aes(y=Y, color="Y"))+
  ylab("Physical processes")+
  scale_color_manual(name="Definitions", values=c("Z"= "red","Y"="blue"))
```

**Answer**
Both variables seem to follow a same overall pattern for example, higher peaks at the beginning then a 
more of a decrease for higher X values. But the similar pattern can be mostly found when X is less than 3,
as they both increase and decrease at around the same time. Although Z has some missing values.
For comparison of the variation, for smaller X values there is a higher variation. The opposite could be found
for higher X values as the variation seems to decrease.


## Question 2.2
**The goal is to derive an EM algorithm that estimates**$\lambda$.

**Answer**
Probability density function of exponential distribution is equal to
$\lambda*e^{-\lambda*x}$

Therefore,
$f(Y_i) = \frac{X_i}{\lambda}*e^{-\frac{X_i*Y_i}{\lambda}}$
$f(Z_i) = \frac{X_i}{2\lambda}*e^{-\frac{X_i*Z_i}{2\lambda}}$

As we are focusing on $\lambda$ for both random variables the jointly probability distribution
can be therefore used. It is important to mention that both random variables are not
dependent on each other as it can be seen in the formula.

Joint probability distribution: 
$f(Y_i,Z_i) = \frac{X_i}{2\lambda}*e^{-\frac{X_i*Z_i}{2\lambda}}*\frac{X_i}{\lambda}*e^{-\frac{X_i*Y_i}{\lambda}}$
= $$\frac{X_{i}^2}{2\lambda^2}*e^{-\frac{X_i*Z_i+2X_i*Y_i}{2\lambda}}$$

Likelihood: $$L(f(Y_i,Z_i)) = (\frac{1}{2\lambda^2})^n* \prod_{i = 1}^{n}X_{i}^2*e^{-\frac{\sum_{i = 1}^{n}X_i*Z_i+2X_i*Y_i}{2\lambda}}$$

Loglikelihood: $$Log (L(f(Y_i,Z_i))) = -nlog(2) - 2nlog(\lambda)+\sum_{i = 1}^{n}log(X_i^2)-\frac{\sum_{i = 1}^{n}X_i*Z_i+2X_i*Y_i}{2\lambda}$$

## Step E
To make this easier to work with 

$-\frac{\sum_{i = 1}^{n}X_i*Z_i+2X_i*Y_i}{2\lambda}$
is split up into $-\frac{\sum_{i = 1}^{n}X_i*Z_i}{2\lambda}-\frac{\sum_{i = 1}^{n}X_i*Y_i }{\lambda}$

Because there are some missing values for $Z_i$ for part $-\frac{\sum_{i = 1}^{n}X_i*Z_i}{2\lambda}$
will be split up into unobserved $-\frac{\sum_{i = r+1}^{n}X_i*Z_i}{2\lambda}$ and observed $-\frac{\sum_{i = 1}^{r}X_i*Z_i}{2\lambda}$.
The unobserved observations have expected value as the random variable.
The expected value of exponential distribution is $\frac{1}{\lambda}$

Therefore, $-\frac{\sum_{i = r+1}^{n}X_i*Z_i}{2\lambda}$ = $-\frac{\sum_{i = r+1}^{n}X_i*\frac{2\lambda_k}{X_i}}{2\lambda}$ = $$-\frac{(n-r)\lambda_k}{\lambda} $$ 

The final formula for **Step E** is:
$$Q(\lambda, \lambda^t) = -nlog(2) - 2nlog(\lambda)+\sum_{i = 1}^{n}log(X_i^2)-\frac{\sum_{i = 1}^{r}X_i*Z_i}{2\lambda}-\frac{(n-r)\lambda_k}{\lambda}-\frac{\sum_{i = 1}^{n}X_i*Y_i }{\lambda}$$

## M step
The partial derivative is taken for $\frac{\partial Q(\lambda, \lambda^t)}{\partial \lambda} = 0$.
$$\frac{-2n}{\lambda}+\frac{\sum_{i = 1}^{r}X_i*Z_i}{2\lambda^2}+\frac{(n-r)\lambda_k}{\lambda^2}+\frac{\sum_{i = 1}^{n}X_i*Y_i }{\lambda^2} = 0$$

$$\lambda = \frac{\sum_{i = 1}^{r}X_i*Z_i+2(n-r)\lambda_k+2\sum_{i = 1}^{n}X_i*Y_i}{4n}$$

## Question 2.3

Implement this algorithm in R, use $\lambda_0$ = 100 and convergence criterion "stop if the change
in $\lambda$ is less than 0:001". What is the optimal $\lambda$ and how many iterations were required to
compute it?
```{r}
#Question 3
EM<-function(x,y,z, kmax){
  Zobs <- z[!is.na(z)]
  Zmiss <- z[is.na(z)]
  Xobs = x[!is.na(z)]
  n <- length(c(Zobs, Zmiss))
  r <- length(Zobs)
  k<-1
  
  eps = 0.001
  prev_lambda = 100
  curr_lambda = (sum(Xobs*Zobs) + 2*(n-r)*prev_lambda + 2*sum(x*y))/(4*n) 
  #For k=1, the current lambda will use the prev_lambda value, as usually
  #this step is done in M-step.
  while ((abs(prev_lambda-curr_lambda)>eps) && (k<(kmax+1))){
    prev_lambda<-curr_lambda
    
    # The E-step is not needed for this case
    # M-step
    curr_lambda = (sum(Xobs*Zobs) + 2*(n-r)*curr_lambda + 2*sum(x*y))/(4*n)
    k<-k+1
  }
  return(c(k,curr_lambda))
}
results = EM(df$X,df$Y,df$Z,100)

```


**Answer**
Optimal $\lambda$ is `r results[2]` and required `r results[1]` iterations.


## Question 2.4
Plot E[Y] and E[Z] versus X in the same plot as Y and Z versus X. Comment whether
the computed $\lambda$ seems to be reasonable.

```{r, echo=FALSE}
df$e_y = results[2]/df$X
df$e_z = (2*results[2])/df$X

#Question 4 
p = ggplot(data=df, aes(x=X))+
  geom_line(aes(y=e_y, color="E(Y)"))+
  geom_line(aes(y=e_z, color="E(Z)"))+  
  geom_line(aes(y=Z, color="Z"))+
  geom_line(aes(y=Y, color="Y"))+
  ylab("Physical processes")+
  scale_color_manual(name="Definitions", values=c("E(Y)"="cyan", 
                                                  "E(Z)"="black",
                                                  "Z"= "red",
                                                  "Y"="blue"))
p
```


**Answer**
The $\lambda$ seems to be reasonable as both variables follow their corresponding expected value.
It can be seen that as X increases, both expected value and the corresponding variable decreases.